### 2024-12-14 background.tex
Added HOL section in Preliminaries (minor changes in Introduction, Background), citation bugs

The Central Limit Theory (CLT) is one of the main theoretical results from the  Probability theory, a bridge from individual randomness to aggregate predictability. It marks the gradual approach of mathematical rigor and upgrading of the theorem used by addressing more and more difficult problems. Over centuries, different forms of CLTs have been derived, each of which has added to its understanding convergence and applications. This chapter will present the historical development of the CLT that describes its milestones and contributions. It will also examine the modern theorem proving tools and ways of formalization, focusing on Lyapunov's approach.

@@

One of the most significant achievements in this area was the formalization of CLT within the Isabelle/HOL proof assistant. This was made possible mainly by the rich library related to complex analysis in Isabelle/HOL, which provided the means to make use of characteristic functions within the proof. Characteristic functions are Fourier transforms of probability distributions that make convergence easier to analyze by drawing on their special mathematical properties. Central to the approach are two key steps: first, pointwise convergence of the characteristic function for the normalized sum of independent random variables to that of the standard normal distribution should be demonstrated; and second, by the application of the Lévy Continuity Theorem, one obtains the desired convergence in distribution. These two steps are actually the core of the proof that eventually leads to the verification of CLT \cite{billingsley2017probability}.

>>>>

The Central Limit Theory is one of the main theoretical results from the  Probability theory, a bridge from individual randomness to aggregate predictability. It marks the gradual approach of mathematical rigor and upgrading of the theorem used by addressing more and more difficult problems. Over centuries, different forms of CLTs have been derived, each of which has added to its understanding convergence and applications. This chapter will present the historical development of the CLT that describes its milestones and contributions. It will also examine the modern theorem proving tools and ways of formalization, focusing on Lyapunov's approach.

@@

One of the most significant achievements in this area was the formalization of CLT within the Isabelle/HOL proof assistant. This was made possible mainly by the rich library related to complex analysis in Isabelle/HOL, which provided the means to make use of characteristic functions within the proof. Characteristic functions are Fourier transforms of probability distributions that make convergence easier to analyze by drawing on their special mathematical properties. Central to the approach are two key steps: first, pointwise convergence of the characteristic function for the normalized sum of independent random variables to that of the standard normal distribution should be demonstrated; and second, by the application of the L\'evy Continuity Theorem, one obtains the desired convergence in distribution. These two steps are actually the core of the proof that eventually leads to the verification of CLT \cite{billingsley2017probability}.

### 2024-12-14 introduction.tex
Added HOL section in Preliminaries (minor changes in Introduction, Background), citation bugs

However, formalizing CLT is far from straightforward. There exist numerous variants of this theorem, together with their numerous proofs, ranging from classical independent and identically distributed (i.i.d.) to some more general versions, such as Lyapunov's theorem \cite{billingsley2017probability}. Each of those proofs uses advanced mathematical tools, including characteristic functions, moment generating functions, or even higher-order moments, according to the situation. These proofs require advanced mathematical infrastructure, which is not available with all proof assistants. In particular, the case of HOL4 \cite{slind2008brief} has found it impossible to use some of the approaches, particularly the one based on characteristic functions, because the strong complex analysis library was lacking.

Instead, this thesis follows another path: formalizing in the HOL4 proof assistant a generalized version of the CLT by Lyapunov's theorem. Lyapunov's approach is particularly appealing, as it has no direct use of either complex analysis or characteristic functions. Instead, it pays attention to moment-based convergence criteria, trying to provide its conditions for convergence using Taylor expansions, Big-O notation of error bounds, and Lyapunov's inequality. This approach aligns well with the existing mathematical infrastructure of HOL4 while simultaneously expanding its capabilities.

This project formalizes Lyapunov's proof in HOL4 and demonstrates the feasibility of verifying advanced probabilistic results in HOL4. More specifically, it has added:
\begin{itemize}
    \item Formally defined Taylor expansions in function bounds and error terms.
    \item Big-O, allowing accurate representations of asymptotic behavior.
    \item Lyapunov's inequality, the key condition for the convergence between the CLT.
\end{itemize}

@@

However, Lyapunov’s method introduced its own set of challenges for formalization in HOL4:
\begin{itemize}
    \item \textbf{Taylor Expansions:} The proof relies heavy on Taylor expansions to approximate test functions and bound errors. These had to be rigorously formalized in HOL4 from first principles.
    \item \textbf{Big-O Notation:} Reasoning about the asymptotic behavior of error terms involved the formalization of Big-O notation, which did not exist in HOL4 until now.
    \item \textbf{Lyapunov's Inequality:} An important part of the proof was the formalization of Lyapunov's inequality, which gives sufficient conditions for convergence. It was carefully adapted to fit into HOL4's logical framework.
\end{itemize}

Following the Lyapunov's approach, with its valuation on real techniques and elementary methods, proved most important: it served to encapsulate the CLT under the constraints of HOL4 and was entirely consistent toward building reusable mathematics libraries. Taylor series developments, Big-O notion, and Lyapunov's inequality are indispensable not only for this proof but also as a basis of further formalizations in the realm of probability theory and elsewhere.

@@

\section{Structure of the Thesis}
The rest of the thesis is organized as follows. Chapter 2 provides background on the Central Limit Theorem, its history, and the setting of Lyapunov's proof. Related work, including other formalizations, using proof assistants like Isabelle/HOL is also covered. Chapter 3 then details the formalization, discussing the challenges faced along with solutions developed, in particular the formalization of Taylor expansions, Big-O notation and Lyapunov's inequality. The key components of the proof, its completion, and its validation of the CLT are discussed in Chapter 4. Chapter 5 presents a discussion of results, and positioning of this work with respect to related efforts on formal verification. Finally, the thesis is concluded in Chapter 6 by summarizing the contributions made by this work and presenting the vision on future directions that probability theory in HOL4 may take.

>>>>

However, formalizing CLT is far from straightforward. There exist numerous variants of this theorem, together with their numerous proofs, ranging from classical independent and identically distributed (i.i.d.) to some more general versions. Notably, Lyapunov's theorem extends the CLT to independent, non-identically distributed (i.n.i.d.) random variables, offering a broader applicability than the classical i.i.d. case \cite{billingsley2017probability}. Each of those proofs uses advanced mathematical tools, including characteristic functions, moment generating functions, or even higher-order moments, according to the situation. These proofs require advanced mathematical infrastructure, which is not available with all proof assistants. In particular, the case of HOL4 \cite{slind2008brief} has found it impossible to use some of the approaches, particularly the one based on characteristic functions, because the strong complex analysis library was lacking.

Instead, this thesis follows another path: leveraging Lyapunov's theorem and the Lindeberg approach, which avoids characteristic functions in favor of real-valued, moment-based techniques. Additionally, it pays attention to moment-based convergence criteria, trying to provide its conditions for convergence using Taylor expansions, Big-O notation of error bounds, and Lyapunov's inequality. This approach aligns well with the existing mathematical infrastructure of HOL4 while simultaneously expanding its capabilities.

@@

However, Linderberg’s method introduced its own set of challenges for formalization in HOL4:
\begin{itemize}
    \item \textbf{Taylor Expansions:} The proof relies heavy on Taylor expansions to approximate test functions and bound errors. These had to be rigorously formalized in HOL4 from first principles.
    \item \textbf{Big-O Notation:} Reasoning about the asymptotic behavior of error terms involved the formalization of Big-O notation, which did not exist in HOL4 until now.
    \item \textbf{Lyapunov's Inequality:} An important part of the proof was the formalization of Lyapunov's inequality, which gives sufficient conditions for convergence. It was carefully adapted to fit into HOL4's logical framework.
\end{itemize}

Following the Linderberg's approach, with its valuation on real techniques and elementary methods, proved most important: it served to encapsulate the CLT under the constraints of HOL4 and was entirely consistent toward building reusable mathematics libraries. Taylor series developments, Big-O notion, and Lyapunov's inequality are indispensable not only for this proof but also as a basis of further formalizations in the realm of probability theory and elsewhere.

@@

\section{Thesis Contributions}
This thesis formalizes Lyapunov's inequality in HOL4, hence establishing a rigorous framework for convergence in probability proofs. In particular, the necessary mathematical theories and lemmas for the proof of CLT: Taylor expansions to bound errors, Big-O notation for asymptotic behavior analysis, and telescoping techniques for incremental replacements. These lemmas are used to formalize the Lyapunov's theorem and Lindeberg's approach, with a subsequent complete proof of CLT within HOL4. Unlike other approaches using characterization or moment-generating functions, this work does not use complex analysis, but works with the methods compatible with the real-valued nature of HOL4.

This thesis, beyond formalizing CLT, enriches libraries for probabilistic analysis that can be reused beyond, such as statistical inference and machine learning. This addresses some foundational gaps in HOL4 and demonstrates the use of HOL4 in advancing formalized mathematics.

\section{Structure of the Thesis}
The rest of the thesis is organized as follows. Chapter 2 provides background on the Central Limit Theorem, its history, and the setting of Lyapunov's proof. Related work, including other formalizations, using proof assistants like Isabelle/HOL is also covered. Chapter 3 then details the preliminaries, including HOL4 as proof engineering and related libraries. The key components of the proof, its completion, and its validation of the CLT are given in Chapter 4. Chapter 5 presents a discussion of results, and positioning of this work with respect to related efforts on formal verification. Finally, the thesis is concluded in Chapter 6 by summarizing the contributions made by this work and presenting the vision on future directions that probability theory in HOL4 may take.

### 2024-12-14 preliminaries.tex
init commit
Added HOL section in Preliminaries (minor changes in Introduction, Background), citation bugs

\chapter{Introduction}
\label{intro}

asdfsdfasdf
\newpage
asdasasfdfdf

% \begin{figure*}
% \centering
% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
% \caption{caption}
% \label{figCh1:figure}
% \end{figure*}

\renewcommand\thefigure{\thechapter .\arabic{figure}}

>>>>

\chapter[Preliminaries]{Preliminaries}
\label{chap:3}

This chapter describes an overview of the theoretical and formal theories required for the formalization of the Central Limit Theorem . This includes HOL Formalization, Measure Theory, Lebesgue Integration, and Probability Theory.
\section{HOL Formalization}
Higher Order Logic (HOL) \cite{hol4, slind2008brief} is derived from the Logic of Computable Functions (LCF) \cite{gordon1979edinburgh, milner1972logic} created by Robin Milner and colleagues in 1972. HOL is an adaptation of Church's Simple Theory of Types (STT) \cite{church1940formulation}, where a higher-order version of Hilbert's choice operator \(\varepsilon \), Axiom of Infinity, and Rank-1 polymorphism have been added. HOL4 implements the original HOL framework, while other theorem provers in the HOL family, such as Isabelle/HOL, include important extensions. Such a simple logical basis makes HOL more accessible than those systems founded on much more advanced dependent type theories, such such the Calculus of Inductive and Co-Inductive Constructions constructed by Coq. Therefore, theories and proofs founded on HOL are easier for a layman to comprehend rather than being lost in a complicated type theory.

HOL refers both to the logical system and the software implementing it. HOL4 is the latest version of this software and written in Standard ML (SML), a general-purpose functional programming language.  SML has played the most vital role in the HOL4 for implementing its core engine, enabled automation due to which proof tactics have been written in that and also for interaction, whether it is through a proof script or in direct correspondence with the user. Integrated SML gives a way in which HOL4 is versatile and can easily be extended such that complex verification tools are provided to develop the management of proofs by a user efficiently.

% HOL logic is based on a formal system of typed logical terms, where types are set-theoretic interpretation sets in a universe \(U\). The HOL type system is, relative to dependent types, considerably less complex yet still suitable for the formalization of various disciplines. HOL comes in with four type categories. As part of its Backus-Naur Form grammar it allows for:
% \begin{itemize}
%     \item \textbf{Primitive types:} For example, \texttt{bool} (representing a two-element set \( \textbf{2} \)) and \texttt{ind} (representing an infinite set \( I \)).
%     \item \textbf{Type constructors:} Operators creating new types, such as functions (`->`) and Cartesian products (`*`).
%     \item \textbf{Type variables:} Enable polymorphism; it is thus possible to write generic theorems and functions.
%     \item \textbf{Defined types:} Derived types, either as subsets or as bijections of basic types.
% \end{itemize}

HOL terms are representatives of such things and the grammar includes constants, variables, applications (function calls) and lambda abstractions. Quantifiers, for instance universal (!x. P(x)) and existential (?x. P(x)), are also provided in HOL - they are defined as specific lambda functions.

\section{Measure Theory}

\section{Lebesgue Integration Theory}

\section{Probability Theory}
